
`\part{Allocation of the kidney}`{=latex}

# Kidney allocation in Eurotransplant {#CHprefacekidney}

Up to 100 million Europeans have Chronic Kidney Disease (CKD) [@Vanholder2021].
CKD is characterized by the progressive loss of kidney function, and its causes
include diabetes, hypertension, and glomerulonephritis. Patients with CKD
can eventually progress to End-Stage Renal Disease (ESRD), which means that they
become dependent on renal replacement therapies (RRTs) such as dialysis or 
kidney transplantation. Whereas other forms of end-stage organ failure cannot
be managed with therapy, end-stage kidney failure can be managed for extended
periods with dialysis [@HaaseKromwijk1999].
However, kidney transplantation is the preferred treatment option for most 
patients with ESRD because it is associated with a reduction in patient mortality 
and improves the patient's quality of life [@Tonelli2011].

Kidney transplantation became the preferred treatment option for patients with
ESRD in the 1980s, when substantial improvements were observed in post-transplant
outcomes [@HaaseKromwijk1999]. As a result, the kidney waiting list in Eurotransplant 
expanded from less than 2,000 patients in 1980 
to approximately 10,500 patients in 1994 [@eurotransplantAR2005]. Since then, the number of patients awaiting kidney transplantation has remained stable in
Eurotransplant with approximately 10,000 patients currently waiting for a kidney transplant.

Eurotransplant allocates kidneys from deceased-organ donors to these waiting patients. 
When Eurotransplant was initiated, candidates were prioritized solely based
on Human Leucocyte Antigen (HLA) matching. Currently, 
Eurotransplant uses an allocation system which tries to balance HLA matching 
with fairness considerations. In this chapter, we describe the history of this 
system and some contemporary challenges in the allocation of deceased-donor kidneys.
We also briefly describe the work included in this thesis that can inform
discussions on addressing these challenges.

## HLA matching versus fairness 

In the late 1960s, immunologists observed that the best outcomes after kidney 
transplantation were observed in pairs of siblings who had compatible ABO blood groups
and matching leukocyte antigen groups [@vanDorp2001]. Anticipating that 
such a matching effect also existed when transplanting kidneys from deceased donors,
Jon van Rood proposed Eurotransplant, a program in which a central office
would use a computer to locate the best-matching candidate for the kidneys that
became available from deceased donors [@vanRoodET1967].
Shortly after Eurotransplant was initiated in 1967, similar organizations were
established in other geographic regions, motivated by the idea that outcomes after
deceased-donor transplantation could only be improved through collaboration [@bakJensenShareNotShare2008].

Already in 1970, this idea became controversial with a strong debate emerging between 
immunologists and clinicians on the actual importance of leukocyte-antigen 
matching in deceased-donor kidney transplantation [@daussetHLAStory1995; @vanRood2004].
The commitment of immunologists to HLA matching was strengthened in the 1970s
and 1980s, as they unraveled many complexities of what became known
as the Human Leucocyte Antigen (HLA) system. The discovery of HLA-DR antigens in 1977 
was particularly important, as these antigens were shown to have a stronger effect
on prognosis after kidney transplantation than the earlier discovered HLA-A
and HLA-B antigens [@Persijn1978; @Persijn1981; @bakJensenShareNotShare2008]. 

At the same time, it was recognized that HLA matching was far from a magic bullet;
many patients with poorly matched kidneys had good outcomes after
transplantation [@terasaki1990history; @vanRood2004]. Based on such observations,
clinicians questioned whether HLA matching was worth the effort, especially because 
the practice resulted in disparities in access to transplantation [@Kjellstrand1988]. 
For example, in the United States, it was observed that black patients were waiting
three times longer than white patients for a kidney transplant.
Thomas Starzl, who is known as "the father of modern transplantation", criticized the blind focus
on HLA matching in kidney allocation, referring to it as _"the institutional 
organization of racial bias"_ [@vanDorp2001, p.159]. To reduce such disparities,
Starzl's transplant center instead used a point system to prioritize candidates
for transplantation, in which tissue matching played a role that was 
_"significant but far from overriding"_ [@starzlMultifactorialSystemEquitable1987].

This criticism of HLA matching intensified in the 1980s with the advent of cyclosporine
A. This immunosuppressant strongly reduced the incidence of acute rejection of kidney
transplants, and was responsible for the major improvements in kidney transplant
outcomes that were observed in the 1980s. Shortly after the introduction of 
cyclosporine A, several single-center studies
suggested that this immunosuppressant had made HLA matching in kidney transplantation 
redundant [@greensteinEvidenceThatZero1990; @harfmannRenalTransplantationUsing1989]. 
These early findings have since been contradicted by multi-center and
registry-based studies, which showed that HLA matching continues to be strongly 
associated with graft and patient survival [@opelzCORRELATIONHLAMATCHING1985; @Opelz2007].
`\newpage`{=latex}
Based on these later studies, consensus has arisen that the best outcomes are observed 
in patients who receive transplants with zero mismatches on the HLA-A, HLA-B,
and HLA-DR loci [@opelzCORRELATIONHLAMATCHING1985; @morrisAnalysisFactorsThat1999]. However, 
the extensive polymorphism of the HLA system means that perfectly matched kidneys
are not available for most candidates. In fact, it was already known in the 1980s
that only about one in five patients can receive a perfectly matched kidney
with a waiting list of 10,000 patients [@CICCIARELLI1987]. Most patients
thus have to settle for a kidney with HLA mismatches. In that case, consensus is
that HLA-DR matching is more important than HLA-B matching, which in turn is more
important than HLA-A matching [@Ting1985; @joyseyTissueTypingPolicy1989; @thorogood1990].

Because of the role of HLA matching was contended, kidney allocation first
became more decentralized in the 1980s [@bakJensenShareNotShare2008]. For example,
in 1988 Eurotransplant introduced a kidney allocation system that mandated 
exchange of kidneys between centers only for (a) patients who had zero HLA mismatches 
with the kidney and (b) the so-called "*hyper-immunized patients*" (see Section \@ref(sec:introam))
[@HaaseKromwijk1999]. Otherwise, the center responsible for procurement of the kidneys 
could choose patients from their own waiting list for transplantation, subject to blood group 
compatibility rules and minimum HLA match criteria (defined as 2 DR matches, or 
1 B + 1 DR match) [@HaaseKromwijk1999]. Procurement centers also had the option to
voluntarily offer the kidneys to the Eurotransplant pool. Candidates would then be prioritized by
compatibility at the HLA-DR locus, followed by compatibility at the HLA-B locus, 
and finally compatibility at the HLA-A locus [@HaaseKromwijk1999].

When Eurotransplant evaluated this system in the 1990s, concerns were raised
over the fact that (i) 10 percent of patients consistently waited more than five years
for kidney transplantation, (ii) transplant rates were low for candidates with
rare HLA types or homozygosity at any of the HLA loci, and (iii) procurement
and transplant rates were imbalanced between the centers and countries 
[@DeMeester1998_booksection]. These fairness concerns 
eventually led to the adoption of the Eurotransplant Kidney Allocation System 
(ETKAS) in 1996, which was based on a point system that had been proposed by
Wujciak and Opelz in the years before. Besides this program, Eurotransplant
has used two other programs to allocate deceased-donor kidneys since the 1990s:
the Acceptable Mismatch (AM) program for the allocation of kidneys to hyper-immunized candidates (Section
\@ref(sec:introam)), and the Eurotransplant Senior Program (ESP) for the allocation
of kidneys from donors aged 65 years or older (Section \@ref(sec:introesp)).

`\newpage`{=latex}

## Kidney allocation programs in Eurotransplant

### The Eurotransplant Kidney Allocation System (ETKAS)

Wujciak and Opelz tried to address the dilemma between HLA
matching and fairness by using computer simulations to develop a point system
for kidney allocation [@wujciakProposalImprovedCadaver1993a; @Opelz1998]. This
system awarded points for the absence of HLA mismatches, but also points for
a candidate's waiting time to ensure that no candidate would be left behind.
Additionally, to avoid extreme waiting times for difficult-to-match patients, 
the system awarded points for the _"mismatch probability"_, a quantification of 
how common favorably matched kidneys are for the patient.

Eurotransplant implemented a modified version of this point system for kidney
allocation in 1996, and the resulting program has become known as the 
Eurotransplant Kidney Allocation System (ETKAS) program. The modifications made
included that ETKAS awards extra priority to pediatric candidates, and implements
a mechanism to balance the international transfer of kidneys [@Opelz1998]. 

One year after the introduction of ETKAS, in 1997, it was concluded that ETKAS
successfully achieved its primary goals; the system had increased the number of transplantations
among pediatric and long-waiting candidates and it had corrected imbalances in 
the international exchange of kidneys [@demeesterNewEurotransplantKidney1998; @Persijn2006].
Despite this apparent success, it was also pointed out by Wujciak and Opelz that
ETKAS _"could not be considered a final and unalterable product"_, which instead 
would require _"continued fine-tuning depending on new developments"_ [@Opelz1998]. 

In light of these comments, it is not surprising that several refinements have 
been made to ETKAS since 1996. One refinement concerns the definition of
waiting time. When ETKAS was initiated, waiting time was defined as the number of days 
a candidate had spent on the waiting list, which conferred an advantage to patients
who were referred early to the kidney waiting list. To address this, Eurotransplant
redefined waiting time to the number of days a candidate had received dialysis
in January 2000 [@DeMeester2000]. A second issue was that ABO
blood group O kidneys could be transplanted in non-O candidates in case of 
a zero mismatch, which led to an increased waiting list mortality and longer 
waiting times for blood group O patients [@glander2010]. In June 2010, this 
issue was addressed by always requiring candidates and donors to have identical
ABO blood groups in ETKAS.

Shortcomings of ETKAS have also been identified for which an adequate solution
has remained elusive. A notable example is that the ETKAS point system places
equal emphasis on the HLA-A, HLA-B and HLA-DR loci, which has been described to lead to 
additional mismatches on the HLA-DR locus 
[@demeesterNewEurotransplantKidney1998; @vereerstraetenExperienceWujciakOpelzAllocation1998]. 
Several proposals have
been made to instead emphasize HLA-DR matching in ETKAS (e.g., Doxiadis et al.  [@doxiadisSimplerEquitableAllocation2007])
which have been rejected by national competent authorities (e.g., Heemskerk et al.  [@heemskerkRegionalKidneyAllocation2009]). 
As a result, ETKAS still places equal emphasis on HLA matching at the A locus
as on matching at the DR locus.

Another example of such a shortcoming are the mismatch
probability points (MMPPs), which are used to facilitate access to transplantation for
difficult-to-match candidates. One issue is that MMPPs are calculated using a
formula that disregards HLA haplotype linkage disequilibrium [@Wujciak1997]. 
A second issue is that at most 100 MMPPs are awarded in ETKAS, which may be
insufficient to provide immunized patients with equality of opportunity
[@susal2015; @ziemannUnacceptableHumanLeucocyte2017; @zecherImpactSensitizationWaiting2022a].

### Immunized candidates \& the Acceptable Mismatch (AM) program {#sec:introam}

Many kidney transplant candidates have developed
antibodies against donor HLA antigens before they are listed for a kidney transplant,
for instance because of a prior blood transfusion, a pregnancy,
or a previous transplantation. If a donor carries the HLA antigens against which the candidate
has developed such donor-specific antibodies (DSAs), transplantation can be
contraindicated because it is likely to result in antibody-mediated rejection 
of the kidney. 

In Eurotransplant, transplant centers have historically been required to conduct
a complement-dependent cytotoxicity (CDC) crossmatch before transplantation, 
in which the serum of the patient is mixed with lymphocytes isolated from the donor. If such a crossmatch
leads to the cell death of the lymphocytes, transplantation is strongly contraindicated
[@patelSignificancePositiveCrossmatch1969], and even prohibited in Germany.
In the past, transplantation 
centers were also required to regularly test the sera of their patients against 
a representative panel of donors to determine their candidates' panel-reactive antibody (PRA) levels.
Such tests are known as the CDC-PRA tests and the obtained PRA levels quantify
the percentage of the donor pool against which the candidate has a positive
crossmatch. CDC testing could also be used to derive
against which antigens a candidate had developed DSAs, albeit at limited resolution.
Since the 1980s, centers have been able to report such antigens to Eurotransplant
as _unacceptable antigens_,
which ensures that there candidates would not be offered kidneys that carry those
antigens.

Having a pre-existing immunization can thus prevent a candidate from receiving a
kidney transplant, in particular for candidates with high PRA levels.
This can lead to an accumulation of immunized patients on the waiting list, which
was indeed observed in the late 1980s. In 1988, Frans Claas and Jon van Rood proposed to 
strategically select donors for the _"hyper-immunized"_ candidates based on _"acceptable antigens"_
[@Claas1988], which formed the basis of the Acceptable Mismatch (AM) program that
was initiated in 1989. These acceptable antigens were defined as antigens against which
an immunized patient had not developed DSAs, and could be identified by crossmatching the candidate's
serum against panels of HLA-typed blood donors that carried exactly one antigen
mismatch with the patient [@Claas1988; @Heidt2015]. In the
AM program, highly immunized candidates (PRA >85%) are given priority
to donors who carried HLA antigens that were either matching with or acceptable to the
candidate. Since its initiation, the AM program has continually been updated by
the Eurotransplant Reference Laboratory (ETRL). For example,
acceptable antigens were initially only defined for HLA-A and HLA-B, while currently 
five HLA loci are used for AM allocation [@ETHist2025].

CDC-PRA assays, in which a candidate's sera is tested against a panel of donor
lymphocytes, has thus had an important role in determining a candidate's degree
of immunization. In recent years this technology has largely been replaced by solid-phase assays,
in which the patient's serum is directly tested against isolated HLA antigens [@Mamode2022].
Centers regularly screen the sera of their candidates with these solid-phase assays and
report the donor-specific antibodies that are identified using solid-phase assays as unacceptable to 
Eurotransplant. Notable is that these solid-phase assays are more sensitive
than CDC-PRA assays, which means that DSAs can be identified that lack
CDC-reactivity. Transplantation in the presence of such DSAs is also associated
with inferior outcomes [@Amico2009], but not an absolute contraindication [@Mulley2011]. 
Instead, European guidelines recommend that the decision to transplant in the
presence of such a DSA should be based on an individualized risk assessment. 
This risk assessment should critically evaluate the DSA in light of potential 
sensitization events and should assess the impact of designating it as unacceptable 
on the likelihood of finding an HLA-compatible kidney [@Mamode2022].

Since 2016, Eurotransplant calculates a _virtual_ PRA (vPRA) for each candidate
based on their reported unacceptable antigens. This vPRA is defined
as the percentage of donors that carry the unacceptable antigens in a database
of 10,000 donors that is maintained by the ETRL. Advantages of the 
vPRA compared to the PRA are (i) that measurement variability associated with the
PRA can be avoided, and (ii) that the vPRA is directly based on unacceptable antigens,
which makes it a better measure of the relative restriction in the 
donor pool faced by immunized candidates. In 2020, the vPRA completely replaced the PRA in
kidney allocation in Eurotransplant. For example, mismatch probability points have
been awarded since 2020 based on the vPRA instead of the PRA, and candidates 
with a PRA below 85% but a vPRA exceeding 85% can enter the AM program under specific conditions 
[@Heidt2015].^[For the AM program, 
unacceptable antigens either require CDC
reactivity, or documentation that links the DSA to a sensitizing event.]

```{=latex}
\vfill
\newpage
\subsection[Eurotransplant Senior Program]{The Eurotransplant Senior Program}
\label{sec:introesp}
```

```{r, results='asis'}
if (knitr::is_html_output()) {
  cat(
    "### The Eurotransplant Senior Program (ESP) {#sec:introesp}"
  )
}

```


In 1999, Eurotransplant initiated the Eurotransplant Senior Program (ESP), which
is used to offer kidneys from donors 65 years or older with priority to candidates
65 years or older. The aim of ESP was (i) to promote the usage of kidneys from
older donors and (ii) to reduce waiting times for candidates 65 years or older [@smitsEvaluationEurotransplantSenior2002]. These aims were achieved by 
prioritizing candidates solely based on their accrued dialysis time and by offering kidneys
only to candidates located in the vicinity of the donor [@Frei2008]. HLA typing
of ESP donors was not performed in order to maximally reduce cold ischemia times.
This meant that transplantations within ESP were done
without consideration of HLA match quality. Because the donor's HLA typing was 
unknown at the time of allocation, immunized candidates were not allowed
to participate in ESP.

Currently, all countries require the HLA typing of ESP donors to be available
before allocation, which has enabled immunized candidates to participate in
ESP. A Eurotransplant paired kidney donation study has demonstrated
that outcomes for patients 65 years or older also improve with HLA-DR matching [@deFijter2023].
Based on these findings, proposals to prioritize HLA-DR matching in ESP have
been made which are scheduled for implementation in 2025. 

## Contemporary challenges in kidney allocation 

Since the introduction of ETKAS and ESP in the 1990s, the landscape of kidney
transplantation has changed substantially. For example, Eurotransplant's donor and patient
populations have aged due to demographic developments, and the immunological
evaluation of transplant candidates has changed with the introduction of
solid-phase assays. Despite this changing landscape,
the core principles of Eurotransplant's kidney allocation systems have remained
largely unchanged. This raises the question of whether these systems adequately
address contemporary challenges in kidney transplantation.

One such challenge concerns the immunized candidates who depend on ETKAS for
access to transplantation. Transplant professionals from Eurotransplant's kidney transplantation 
centers have regularly expressed concerns that this patient group is disadvantaged 
in ETKAS. Empirical evidence for such a disadvantage comes from Germany where retrospective analyses have
shown that patients with vPRAs exceeding 85% are disadvantaged 
[@ziemannUnacceptableHumanLeucocyte2017; @zecherImpactSensitizationWaiting2022a]. 
In Chapter \@ref(CHvpra), we examine the relation between the vPRA and the
transplant rate using retrospective data from all Eurotransplant member
countries. We indeed find that immunized candidates face reduced access to 
transplantation in ETKAS. European guidelines already recommend that this disparity should be addressed 
by directly awarding points for the vPRA using a sliding scale [@Mamode2022].
In Chapter \@ref(CHetkidneysimulator), we design such a sliding scale for ETKAS.

A second challenge comes from the changing patient and donor demographics. 
Since the Eurotransplant Senior Program (ESP) was introduced in 1999, the
percentage of patients aged 65 years or older at listing has increased from 6% 
in 1999 to 22% in 2024, and the percentage of donors aged 65 years or older has increased
from 11% to 24%. With this changing demographic, concerns have been voiced over the fairness of
ESP. For example, in Germany, the median dialysis time at transplantation 
currently differs by more than four years between candidates aged over 65 and
those under 65 [@Kolbrink2024]. This means that a German candidate who starts
dialysis at age 60 is unlikely to receive a kidney transplant until their 65th
birthday, and is expected to be transplanted shortly thereafter.

Eurotransplant also has not implemented any form of candidate-donor age matching
within ETKAS, with the exception of pediatric candidates, who have priority to
kidneys procured from pediatric donors. Organ allocation organizations in France, 
the U.K. and the U.S. have all implemented mechanisms which direct kidneys from 
young donors to young patients [@Audry2022; @watson2020overview; @israniNewNationalAllocation2014]. 
Calls have been made in the recent literature to also introduce
such continuous candidate-donor age matching in Eurotransplant [@vonsamson-himmelstjernaContinuousDonorrecipientAge2024].
In Chapter \@ref(CHetkidneysimulator), we evaluate a form of continuous-donor
age matching for ETKAS.

A third challenge is that ETKAS has continued to place equal emphasis on HLA-A, 
HLA-B, and HLA-DR matching, while virtually all other organ exchange organizations
emphasize matching at the HLA-DR locus, or matching at both the HLA-B and HLA-DR loci
[@Wu2017]. In Chapter \@ref(CHetkidneysimulator), we examine how waiting list
outcomes are affected if more priority is given to matching at the HLA-B
and HLA-DR loci.

This thesis explores several directions in which kidney allocation in
Eurotransplant could be improved. Instrumental to studying the impact of
these policy changes is
the ETKidney simulator, a discrete-event simulator that mimics ETKAS and
ESP allocation based on data from the Eurotransplant database. We describe and validate this simulator
in Chapter \@ref(CHetkidneysimulator).

```{r eval=FALSE, include=FALSE}

## Miscellaneous



@DeMeester2000: invited comment on issues with ETKAS.


Opelz and Wujciak: What to expect from a good allocation system

- "The new allocation system cannot be considered a final and unalterable
product. Rather, it will require continued fine-tuning depending on new
developments, or the emergence of inequalities that have gone undetected
during the first 2 years"

- An issue that cannot be resolved by computer simulations is the weight of
priorities that should be assigned to various conditions. (...) Organ transplantation is undergoing continuous development, and the significance of risk factors may change in the future, just as it has in the past. Past exclusion criteria were different from the present; and they are likely to change
again in the future. It will be the task of expert committees (including not only
transplant doctors but also representatives of patient associations, ethicists,
social workers, etc.) to determine the importance that should be attached to
these factors in the organ allocation procedure.



What to expect from a good kidney allocation system

- de Meester 1998:

 It should be stressed that any introduction of a change must be preceded by a computer simulation study, to test whether the modification will attain the desired effect and which immediate impact it will have on the other patients on the waiting list-because there is a fixed supply-and thus which alternative is the better one

- Wujciak-Opelz: 

If the highest success rate were the ultimate goal, patients with
serious complicating health conditions (e.g. diabetics), those acutely in need of
transplantation because of rapidly deteriorating health conditions, or highly
sensitized patients would be excluded because experience shows that one cannot
expect an optimal success rate in these patients. Likewise, small children or
patients of advanced age would not be selected because relatively high graft
failure rates have been reported. If, on the other hand, the principle of equitable
access were followed as the sole guideline, patients might receive organs that
are unsuitable for them, e.g. a kidney might be completely histoincompatible
or even crossmatch-positive. It follows that compromises have to be reached,
and that rules have to be established that can be applied uniformly.


Gjertson: proposal on HLA matching for U.S. https://www.nejm.org/doi/full/10.1056/NEJM199104113241505

Cecka. Review on HLA matching: https://onlinelibrary.wiley.com/doi/10.1111/j.1744-313X.2010.00957.x

Since 2023, transplant centers 
can report unacceptable antigens on 11 HLA loci, and donor procurement centers 
are required to support HLA typings electronically to Eurotransplant for these 
11 loci. This has made it possible to implement a virtual crossmatch in
kidney allocation in 2023 [@Heidt2024]. Centers are currently still required to
conduct a physical crossmatch for all kidney transplantations, which may be done
retrospectively for non-immunized candidates [@ETHist2025].

```
