\documentclass[12pt]{article}

% Set page size
\usepackage[paperwidth=7.76in, paperheight=11.01in, margin=0.75in]{geometry}

% Better font and formatting
\usepackage{lmodern}
\usepackage[T1]{fontenc}
\usepackage{parskip}
\usepackage{enumitem}
\usepackage{titlesec}
\usepackage{graphicx}

% Use PT serif font.
\usepackage{fontspec}
\setmainfont{PT Sans}

% Title style
\titleformat{\section}
{\normalfont\Large\bfseries}{\thesection}{1em}{}

% No page numbers
\pagestyle{empty}

\begin{document}
	
	\begin{center}
		{\Large \textbf{STATEMENTS}}\\[1em]
		\setmainfont{PT Serif}
		{\large associated with the thesis}\\[1em]
		{\large \textbf{Organ allocation models in Eurotransplant}}\\[0.5em]
	\end{center}
	
	\vspace{1em}
		
	\begin{enumerate}[leftmargin=1cm, label=\textbf{\arabic*.}]
		\item MELD remains an excellent basis for the allocation of the liver.
		\item If the HLA matching scheme of ETKAS were proposed as a new policy today, it would almost certainly be rejected by national competent authorities.
		\item Eurotransplant is a knowledge-sharing network for transplant centers as much as it is an organ-sharing network.
		\item Ranking transplant candidates by their counterfactual outcomes cannot deliver equality of opportunity, because whether there is equality of opportunity is determined by factual and not hypothetical outcomes.
		\item Most standard assumptions used in causal inference are invalid in the context of a transplant waiting list. (\textit{technical-intrinsic})
		\item The Cox proportional hazards model offers many flexibilities, which are underutilized by both the medical and machine learning research communities.  (\textit{technical-intrinsic})
		\item Many scientific debates could be resolved if experts from different fields -- and those with different roles -- talked more with each other. (\textit{technical-intrinsic})
		\item Science and its funding overemphasize the use of novel techniques and novel ideas. (\textit{technical-intrinsic})
		\item A major barrier to improving complex systems -- of which organ allocation systems are just an example -- is the lack of clarity on who is behind the steering wheel.  (\textit{general social})
		\item It is unfair to ask organizations to `make things fair' without providing a definition of fairness.  (\textit{general social})


	\end{enumerate}
	
	\vfill
	
	\newpage
	
		\begin{center}
			{\Large \textbf{STELLINGEN}}\\[1em]
			\setmainfont{PT Serif}
			{\large associated with the thesis}\\[1em]
			{\large \textbf{Organ allocation models in Eurotransplant}}\\[0.5em]
		\end{center}
		\vspace{1em}
	
		\begin{enumerate}[leftmargin=1cm, label=\textbf{\arabic*.}]
		\item MELD is nog steeds een uitstekende basis voor de leverallocatie.
		\item Als het HLA-matchingssysteem van ETKAS vandaag als nieuw beleid zou worden voorgesteld, zou het vrijwel zeker afgewezen worden door de nationale autoriteiten.	
		\item Eurotransplant draait net zo goed om de uitwisseling van kennis tussen de transplantatiecentra als om de uitwisseling van organen.	
		\item Het toewijzen van organen enkel op basis van de \textit{counterfactual outcomes} van de patiënt staat op gespannen voet met kansengelijkheid, omdat kansengelijkheid draait om de daadwerkelijke uitkomsten van patiënten en niet om de hypothetische.
		\item De meeste standaardaannames die gebruikt worden in de \textit{causal inference} zijn problematisch als het gaat om de transplantatiewachtlijst. (\textit{technisch-wetenschappelijk})
		\item Het Cox proportional hazards model biedt een grote mate van flexibiliteit, die in zowel medisch als machine learning onderzoek onderbenut wordt.\\ (\textit{technisch-wetenschappelijk})
		\item Veel wetenschappelijke discussies zouden opgelost kunnen worden als deskundigen uit verschillende vakgebieden -- of deskundigen met verschillende rollen -- meer met elkaar in gesprek gingen. (\textit{technisch-wetenschappelijk})
		\item De wetenschap en wetenschapsfinanciering leggen te veel nadruk op het gebruik van nieuwe technieken en ideeën. (\textit{technisch-wetenschappelijk})
		\item Een belangrijk obstakel voor de verbetering van complexe system -- waaronder de allocatie van organen valt -- is de onduidelijkheid die er bestaat over wie er aan de knoppen draait.	(\textit{sociaal-maatschappelijk)})
		\item Het is oneerlijk organisaties de opdracht te geven om beleid ``eerlijk" te maken, zonder te definiëren wat onder eerlijkheid verstaan moet worden.\\ (\textit{sociaal-maatschappelijk})

	\end{enumerate}
	\vfill
	\begin{center}
		%\rule{0.5\linewidth}{0.5pt}\\
		{\small Hans de Ferrante}
	\end{center}
	
	
\end{document}
